---
title: "Activity Analysis"
author: "Chris Njuguna"
date: "July 21, 2016"
output: 
  html_document: 
    keep_md: yes
    self_contained: no
---

###Read in the data and convert the dates from characters to date objects

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='figure/',
                      echo=TRUE, warning=FALSE, message=FALSE, cache=FALSE, base64_images=FALSE)

#Load the data
activity <- read.csv("activity.csv", sep=",", na.strings = "NA", header = TRUE)

#convert date strings into dates
activity$date <- as.Date(activity$date)

```

###What is mean total number of steps taken per day?

```{r}
#calculate the mean number of steps taken for each day
mean_daily_steps <- tapply(X=activity$steps, INDEX = activity$date, FUN=mean, na.rm=TRUE)

```

A histogram of the number of steps taken each day is shown below

```{r}
hist(mean_daily_steps)

```

The mean number of total steps per day is `r mean(mean_daily_steps, na.rm=TRUE)` while the median is `r median(mean_daily_steps, na.rm=TRUE)`

###What is the average daily activity pattern?

A timeseries plot of the average of the 5-minute intervals across all days

```{r}
mean_5min_intervals <- tapply(X=activity$steps, INDEX = activity$interval, FUN=mean, na.rm=TRUE)
plot(seq(0, 1435, by = 5), mean_5min_intervals, type="l", xlab = "5 min intervals", ylab = "Mean steps")

#calculate the max steps
max_mean_5min_intervals <- max(mean_5min_intervals,na.rm=TRUE)

#find the index with the highest number of steps
idx_max_mean_5min_intervals <- which(mean_5min_intervals == max_mean_5min_intervals)
max_intvl <- idx_max_mean_5min_intervals*5
```
The 5 min interval with the highest number of steps is `r max_intvl` with `r max_mean_5min_intervals` steps.

###Imputing missing values

```{r}
num_NA <- sum(is.na(activity))

#calculate the mean across all steps
mean_all_intervals <- mean(activity$steps, na.rm=TRUE)

#copy the activity dataset
activity_noNA <- activity

#replace all NAs with the mean across all steps
activity_noNA[is.na(activity_noNA$steps),"steps"] <- mean_all_intervals

mean_5min_intervals_noNA <- tapply(X=activity_noNA$steps, INDEX = activity_noNA$interval, FUN=mean, na.rm=TRUE)

```

A histogram of the number of steps taken each day with imputed data is shown below

```{r}

hist(mean_5min_intervals_noNA)

```

There is a great change in the distribution of the steps. The imputed data displays a strong right skew while the data with NAs is almost normally distributed.

The mean number of total steps per day is `r mean(mean_5min_intervals_noNA, na.rm=TRUE)` while the median is `r median(mean_5min_intervals_noNA, na.rm=TRUE)`

###Are there differences in activity patterns between weekdays and weekends?

```{r}
activity$wkday <- sapply(weekdays(activity$date, abbreviate = TRUE), FUN = function(x) if (x == "Sat" | x == "Sun") {"weekend"} else {"weekday"})

activity$wkday <- as.factor(activity$wkday)

wkdays <- which(activity$wkday == "weekday")
wkends <- which(activity$wkday == "weekend")


wkdays <- which(activity$wkday == "weekday")
wkends <- which(activity$wkday == "weekend")

weekday_means <- tapply(X=activity[wkdays,]$steps, INDEX = activity[wkdays,]$interval, FUN=mean, na.rm=TRUE)

weekend_means <- tapply(X=activity[wkends,]$steps, INDEX = activity[wkends,]$interval, FUN=mean, na.rm=TRUE)

par(mfrow=c(1,2))

rng <- range(weekday_means, weekend_means)

plot(seq(0, 1435, by = 5), weekday_means, type="l", ylim=rng, xlab = "5 min intervals", ylab = "Mean steps", sub = "weekday")

plot(seq(0, 1435, by = 5), weekend_means, type="l", ylim=rng, xlab = "5 min intervals", ylab = "Mean steps", sub = "weekend")

title(main="Average steps per interval", outer = TRUE)

```

There are differences between weekday and weekend activity. There is a slower start to activity on the weekend which culminates in higher volumes in the middle of the day than weekdays. Weekdays however have a higher maximum number of steps early in the day.
