---
title: "Activity Analysis"
author: "Chris Njuguna"
date: "July 21, 2016"
output: 
  html_document: 
    keep_md: yes
    self_contained: no
---

###Read in the data and convert the dates from characters to date objects

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='figure/',
                      echo=TRUE, warning=FALSE, message=FALSE, cache=FALSE, base64_images=FALSE)

setwd("/media/NewVolume/Coursera/JHUDataScienceSpecialization/05_ReproducibleResearch/activity/datasharing")

#Load the data
activity <- read.csv("activity.csv", sep=",", na.strings = "NA", header = TRUE)

#convert date strings into dates
activity$date <- as.Date(activity$date)
```

###What is mean total number of steps taken per day?

```{r}
#calculate the mean number of steps taken for each day
#mean_daily_steps <- tapply(X=activity$steps, INDEX = activity$date, FUN=mean, na.rm=TRUE)
mean_daily_steps <- aggregate(steps ~ date, data=activity, sum, na.rm=TRUE)
```

A histogram of the number of steps taken each day is shown below

```{r}
hist(mean_daily_steps$steps, xlab = "Total daily steps taken")
```

The mean number of total steps per day is `r as.integer(mean(mean_daily_steps$steps, na.rm=TRUE))` while the median is `r as.integer(median(mean_daily_steps$steps, na.rm=TRUE))`

###What is the average daily activity pattern?

A timeseries plot of the average of the 5-minute intervals across all days

```{r}
#mean_5min_intervals <- tapply(X=activity$steps, INDEX = activity$interval, FUN=mean, na.rm=TRUE)
#plot(seq(0, 1435, by = 5), mean_5min_intervals, type="l", xlab = "5 min intervals", ylab = "Mean steps")

mean_5min_intervals <- aggregate(steps ~ interval, data = activity, FUN = mean, na.rm=TRUE)

plot(steps ~ interval, data = mean_5min_intervals, type="l")

#calculate the max steps
max_mean_5min_intervals <- max(mean_5min_intervals$steps,na.rm=TRUE)

#find the index with the highest number of steps
idx_max_mean_5min_intervals <- which(mean_5min_intervals$steps == max_mean_5min_intervals)
max_intvl <- mean_5min_intervals[idx_max_mean_5min_intervals,]$interval

abline(v=max_intvl, col="red", lwd=2)
text(x=max_intvl+600,y=max_mean_5min_intervals, paste("intvl:",max_intvl, "steps:", round(max_mean_5min_intervals), sep=" "))
```

The 5 min interval with the highest number of steps is `r max_intvl` with `r max_mean_5min_intervals` steps.

###Imputing missing values

```{r}
num_NA <- sum(is.na(activity$steps))

#calculate the mean steps for each interval
mean_all_intervals <- aggregate(steps ~ interval, data=activity, FUN=mean, na.rm=TRUE)

#copy the activity dataset
activity_noNA <- activity

intervals_NA <- activity_noNA[is.na(activity_noNA$steps),"interval"]

#replace all NAs with the mean for that 5 min interval
 activity_noNA[is.na(activity_noNA$steps),"steps"]  <- sapply(intervals_NA, FUN = function(x) { mean_all_intervals[mean_all_intervals$interval == x, "steps"]})

#mean_5min_intervals_noNA <- tapply(X=activity_noNA$steps, INDEX = activity_noNA$interval, FUN=mean, na.rm=TRUE)

mean_5min_intervals_noNA <- aggregate(steps ~ date, data=activity_noNA, FUN=sum, na.rm=TRUE)
```

A histogram of the number of steps taken each day with imputed data is shown below

```{r}

hist(mean_5min_intervals_noNA$steps, xlab = "Total steps taken")
```

There is a slight change in that there are higher counts around the center of the histogram.

The mean number of total steps per day is `r as.integer(mean(mean_5min_intervals_noNA$steps, na.rm=TRUE))` while the median is `r as.integer(median(mean_5min_intervals_noNA$steps, na.rm=TRUE))`

###Are there differences in activity patterns between weekdays and weekends?

```{r}
activity$wkday <- sapply(weekdays(activity$date, abbreviate = TRUE), FUN = function(x) if (x == "Sat" | x == "Sun") {"weekend"} else {"weekday"})

activity$wkday <- as.factor(activity$wkday)

#wkdays <- which(activity$wkday == "weekday")
#wkends <- which(activity$wkday == "weekend")

#weekday_means <- tapply(X=activity[wkdays,]$steps, INDEX = activity[wkdays,]$interval, FUN=mean, na.rm=TRUE)

#weekend_means <- tapply(X=activity[wkends,]$steps, INDEX = activity[wkends,]$interval, FUN=mean, na.rm=TRUE)

#par(mfrow=c(1,2))

#rng <- range(weekday_means, weekend_means)

#plot(seq(0, 1435, by = 5), weekday_means, type="l", ylim=rng, xlab = "5 min intervals", ylab = "Mean steps", sub = "weekday")
#
#plot(seq(0, 1435, by = 5), weekend_means, type="l", ylim=rng, xlab = "5 min intervals", ylab = "Mean steps", sub = "weekend")
#
#title(main="Average steps per interval", outer = TRUE)

par(mfrow=c(1,1))

#means <- aggregate(activity$steps, by=list(activity$wkday, activity$interval), mean, na.rm=TRUE)

means <- aggregate(steps ~ interval + wkday, data = activity, mean, na.rm=TRUE)

lattice::xyplot(steps ~ interval | as.factor(means$wkday), data = means, type="l")
```

There are differences between weekday and weekend activity. There is a slower start to activity on the weekend which culminates in higher volumes in the middle of the day than weekdays. Weekdays however have a higher maximum number of steps early in the day.
